---
title: "**Final Exam -- Part 2**"
author: "Jasmeet Singh Saini - 0758054"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Concerned Coffee Connaisseur

## A pretentious coffee snob that you know loves to extol the virtues of blooming coffee grounds when making a cup of coffee. Your friend says that anyone can tell the difference between a bloomed cup of coffee and an unbloomed cup of coffee. To test this, you enlist 20 of your closest coffee drinking friends. For each friend, you make 2 cups of coffee: 1 bloomed and 1 unbloomed, but keep the brewing process identical otherwise. You randomize the order in which the drinker tries the coffee, and ask them for an "enjoyment" score between 1 and 10, including decimals. The data you collect is in the below table, which can be loaded using the code below the table:

```{r}
bloomed <- c(9.2, 8, 9.1, 7.7, 8.1, 7.8, 7, 9.2
             , 8.3, 7, 8.8, 7, 7.4, 8.5, 7.8, 7.1
             , 6.4, 8, 8.4, 7.4)
unbloomed <- c(7.9, 7.9, 7.7, 7.9, 7.7, 7.6, 7.8
               , 7.8, 7.8, 7.9, 7.7, 7.9, 7.8, 7.8
               , 7.8, 7.8, 7.9, 7.8, 7.8, 7.9)

coffee <- data.frame(participantID = 1:20
                     , bloomed = bloomed
                     , unbloomed = unbloomed)
```

### a. What test should you use to determine if bloomed coffee is more enjoyable? (Be as specific as possible.)
  
A **Paired t-test** will be use to determine if bloomed coffee is more enjoyable

### b. What are the degrees of freedom of the test statistic associated with (a)?

The degrees of freedom of the test statistic associated with paired t-test is **19**. The Paired t-test is performed in part c of this question.

### c. Test whether bloomed coffee enjoyment is higher than unbloomed coffee.

## Hypothesis

The $H_{0}$ is Null hypothesis and $H_{A}$ is Alternative hypothesis.

$$
\begin{aligned}
H_{0}:\ &\text{ Bloomed coffee is equally enjoyable as unbloomed coffee} \\
&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \text{vs.} \\
H_{A}:\ &\text{ Bloomed coffee enjoyment is higher than unbloomed coffee}
\end{aligned}
$$

### Test Statistic and _p_-value

Preforming the **Paired t-test*:

```{r}
coffee_t_test <- t.test(coffee$bloomed, coffee$unbloomed, paired = TRUE, alternative = "greater")
coffee_t_test
```

Here, we get the test statistic as **0.53213** and *p*-value as **0.3004**. Also, the degrees of freedom of the test statistic associated with paired t-test is **19**.


### Statistical Decision

We **fail to reject null hypothesis** $H_{0}$ as $p$-value is greater than $\alpha$ level of significance.(i.e., *p*-value > 0.05)

### Conclusion

There is enough evidence to support the null hypothesis. Therefore, we conclude that the statement, "Bloomed coffee is equally enjoyable as unbloomed coffee", is accurate.


### d. Is your friend a pretentious snob, or might there be something to this whole "blooming" thing?

My friend may be a pretentious snob as the result of the paired t-test suggest that "Bloomed coffee is equally enjoyable as unbloomed coffee".
