---
title: "**Final Exam -- Part 2**"
author: "Jasmeet Singh Saini - 0758054"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cautious Caretakers of Climate Control

**Note: the table and data.frame code do NOT match. Please use the data.frame code, regardless.**

## Trent caretakers measure the temperatures in four of the university's buildings at different times of the day and want to know what affects the temperature: the building, the time of day, or the interaction between the two. The data they collected is:

## And can be loaded into a data.frame in R using:

```{r}
btemp <- data.frame(temperature = c(21, 20.3, 20.7, 21.9, 20.7, 21.2
                                    , 20, 19.5, 20.1, 19.2, 19.3, 19.1
                                    , 20.2, 19.8, 20.1, 20.5, 19.8, 20.1
                                    , 20.1, 19.4, 19.9, 19.0, 21.2, 19.5
                                    , 18.9, 19.1, 19.3, 20.1, 20.7, 21.5
                                    , 18.7, 18.9, 19.3,  18.0, 17.9, 19.2)
                    , building = factor(rep(rep(c("DNA", "ENW", "OCA", "TSC")
                                                , each = 3), 3))
                    , time = factor(rep(c("Morning", "Afternoon", "Evening"), 12)
                                    , levels = c("Morning", "Afternoon", "Evening")))
```


### a. What type of test should be used?

A **Two-way ANOVA (analysis of variance) test** should be used. 


### b. Construct an interaction plot. What does this plot indicate? 

To construct an interaction plot in R, we can use the interaction.plot() function:
```{r}

interaction.plot(x.factor     = btemp$building, 
                 trace.factor = btemp$time,
                 response     = btemp$temperature,
                 main         = "Interaction Plot", 
                 xlab         = "University's Building", 
                 ylab         = "Teamperature (degree C)",
                 type         = "b",
                 col          = c("blue", "green", "red")) 
```

We can see from the interaction plot that a changes in the average temperature in the TSC building is observed due to an interaction between the afternoon time with evening and morning time.


### c. Conduct a test to determine if there is an interaction effect on temperature due to building and time of day.

## Hypothesis

The $H_{0}$ is Null hypothesis and $H_{A}$ is Alternative hypothesis.

$$
\begin{aligned}
H_{0}:\ &\text{ There is no interaction effect on temperature due to building and time of day} \\
&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \text{vs.} \\
H_{A}:\ &\text{ There is no interaction effect on temperature due to building and time of day}
\end{aligned}
$$

### Test Statistic and _p_-value

Preforming the *Two-way ANOVA test*:

```{r}
trent_uni <- aov(temperature ~ building * time, data = btemp)
# Getting the summary
summary(trent_uni)
```

Here, we get the test statistic as **0.361** and *p*-value as **0.89605**.

### Statistical Decision

We **accept** the Null Hypothesis $H_{0}$ as $p$-value is greater than $\alpha$ level of significance.

### Conclusion

There is enough evidence to support the null hypothesis. Therefore, we conclude that the statement, "there is no interaction effect on temperature due to building and time of day", is accurate.

### d. Determine if there are any main effects: interpret these main effects (account for the interaction if necessary!). You don't need the full hypothesis testing framework here.

Determining if there are any main effects:

```{r}
main_effect <- lm(temperature ~ building + time, data = btemp)
summary(aov(main_effect))
```

The *p*-value for building is **0.000939**, which is less than *0.05*. And the *p*-value of time is **0.641666** which is greater than *0.05*. Hence, building affects the average temperature, while time does not.


### e. Which building (statistically), if any, is the hottest on average?

To check which is the hottest building on average, we will use Tukey HSD - Statistical Test for Differences.   
```{r}
TukeyHSD(trent_uni, "building")
```

Hence, ENW building is the hottest on average.



