

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Google Play Store Dataset Analysis

The ever-changing mobile landscape is a challenging space to navigate. The percentage of mobile over desktop is only increasing.
Android holds about 53.2% of the smartphone market, while iOS is 43%. To get more people to download your app, you need to make sure they can easily find your app.
Mobile app analytics is a great way to understand the existing strategy to drive growth and retention of future user.
With million of apps around nowadays, the following data set has become very key to getting top trending apps in iOS app store.
This data set contains more than 7000 Apple iOS mobile application details. The data was extracted from the iTunes Search API at the Apple Inc website.

```{r , warning=FALSE,message=FALSE}
# list.files("../input")
library(tidyverse)
library(plyr)
library(dplyr)
library(magrittr)

gplay <- read.csv("googleplaystore.csv")

```

## Table of Contents

>Dataset Description
>Design
>Implementation
>Observations and Insights
```{r, echo=FALSE}
head(gplay)
print(paste("Dimensisons of Dataset are ",dim(gplay)))
```
##Dataset Description

####App:                      Application name
####Category:                 Category the app belongs to
####Rating:                   Overall user rating of the app (as when scraped)
####Reviews:                  Number of user reviews for the app (as when scraped)
####Size:                     Size of the app (as when scraped)
####Installs:                 Number of user downloads/installs for the app (as when scraped)
####Type:                     Paid or Free
####Price:                    Price of the app (as when scraped)
####Content Rating:           Age group the app is targeted at - Children / Mature 21+ / Adult
####Genres:                   An app can belong to multiple genres (apart from its main category). For eg, a                                   musical family game will belong to Music, Game, Family genres.
####Last Updated:             Date when the app was last updated on Play Store (as when scraped)
####Current Ver:              Current version of the app available on Play Store (as when scraped)
####Android Ver:              Min required Android version (as when scraped)

```{r pressure, echo=FALSE,message=FALSE,warning=FALSE}


gplay$Last.Updated[gplay$Last.Updated=="1.0.19 April 1"]= NULL
gplay$Last.Updated <- as.character(gplay$Last.Updated)
gplay$Last.Updated <- as.Date(gplay$Last.Updated,format="%B %d, %Y",tryFormats = "%Y %m %B")

gplay$Rating <- as.numeric(gplay$Rating)
gplay$Reviews <- as.numeric(gplay$Reviews)

gplay$Installs <- gsub(",", "", gsub("\\.", "", gplay$Installs))
gplay$Installs <- as.character(gplay$Installs)
gplay$Installs = substr(gplay$Installs,1,nchar(gplay$Installs)-1)
gplay$Installs <- as.numeric(gplay$Installs)

gplay$Size <- gsub(",", "", gsub("\\.", "", gplay$Size))
gplay$Size <- as.character(gplay$Size)
gplay$Size = substr(gplay$Size,1,nchar(gplay$Size)-1)
gplay$Size <- as.numeric(gplay$Size)


gplay$App <- as.character(gplay$App)

gplay$Category = as.character(gplay$Category)
gplay$Category[gplay$Category=="1.9"]<-NA
gplay$Category <- as.factor(gplay$Category)

gplay$Type = as.character(gplay$Type)
gplay$Type[gplay$Type=="0"]<-NA
gplay$Type[gplay$Type=="NaN"] <- NA
gplay$Type <- as.factor(gplay$Type)

gplay$Price <- as.character(gplay$Price)
gplay$Price = as.numeric(gsub("\\$", "", gplay$Price))

gplay$Genres = as.character(gplay$Genres)


```

###Number of Apps Updated Last Month

```{r, echo=FALSE}

gplay$Month <- format(gplay$Last.Updated, "%b")
gplay$Month <- factor(gplay$Month, labels=c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'))

table(gplay$Month)
```

###Analysis of the Types of Apps and their average User Rating

```{r, echo=FALSE}
##Let us analyze the type of Apps and their average rating
gplay<- na.omit(gplay)
cat = gplay %>% group_by(Category) %>% select(c(App, Category, Rating))
cat = as.data.frame(cat)
str(cat$Rating)

table(cat$Rating)
```

###Analyzing the Type of Apps and their Average Rating 

```{r,echo=FALSE}
#Analyzing the Type of Apps and their Average Rating 
type<- gplay %>% group_by(Type) %>% select(c(Rating,Type, Installs, Category, Price))
type=as.data.frame(type)

table(gplay$Type)
```

###Installations by Type and Categories

```{r,echo=FALSE}
ggplot(type, aes(x=type$Type, y=type$Installs, fill=type$Type))+geom_bar(stat="identity")+labs(x="Type",y="Installs",fill="Types",title="Installations by Type of Apps")

ggplot(type, aes(x=type$Category, y=type$Installs, fill=type$Category))+ geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 90, vjust=0.5))+ labs(title="Installations by Category of Apps",x="Categories",y="Installs",fill="Categories")
```

###Since Game has more installations Lets see the most installed Apps under it

```{r,echo=FALSE}
gp1<- subset(gplay, Category=="GAME", select = c(App, Installs, Rating))
gp1<-top_n(gp1, 10)
ggplot(gp1, aes(x=gp1$App, y=gp1$Installs, fill=gp1$Rating))+geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 90, vjust=0.5))+ labs(title="Most installed Apps under Game Category",subtitle="(All rated above 5)",x="Apps",y="Installs",fill="Rating")
```

###Most Family Apps installed

```{r,echo=FALSE}
gp2<- subset(gplay, Category=="FAMILY", select = c(App, Installs, Rating))
gp2<-top_n(gp2, 10)
ggplot(gp2, aes(x=gp2$App, y=gp2$Installs, fill=gp2$Rating))+geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 90, vjust=0.5))+ labs(title="Most installed Apps under Family Category",subtitle="(All rated above 5)",x="Apps",y="Installs",fill="Rating")
```

###Lets see how are the App's Priced?

```{r,echo=FALSE}
ggplot(gplay, aes(x=gplay$Price))+geom_density(linetype="dashed", color="Black", fill="blue")+ theme(axis.text.x = element_text(angle = 90, vjust=0.5))+labs(title="Lets see how are the App's Priced?",x="App Prices")

paid_data<- subset(gplay, Type=="Paid", select = c(App, Price, Rating,Installs))
ggplot(paid_data, aes(x=paid_data$Installs))+geom_density(color="black",fill="blue")+labs(title="Installation density of Paid Apps",x="Installs")
```

###Let's Analyze content rating and Installs

```{r,echo=FALSE}
ggplot(gplay, aes(x=gplay$Content.Rating, y=gplay$Installs, fill=gplay$Installs))+geom_bar(stat="identity")+labs(title="Content Rating and Installs",subtitle= "Analysis",x="Content Rating",y="Installs",fill="Installs")
```

### Check which category has the highest rating

```{r,echo=FALSE}
ggplot(gplay, aes(x=Rating, y=Category)) +
  geom_segment(aes(yend=Category), xend=0, colour="grey50") +
  geom_point(size=1, aes(colour=Type)) +
  scale_colour_brewer(palette="Set1", limits=c("Free", "Paid"), guide=FALSE) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank()) +
  facet_grid(Type ~ ., scales="free_y", space="free_y") +
  ggtitle("Checking which Category has the highest Rating between Free and Paid Apps")
```

### Number of Ratings

```{r,echo=FALSE}
ggplot(gplay, aes(x= Category, y= Rating, fill = Type)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  ggtitle("Number Of App Ratings Based On Category and Type")
```

###Number of Reviews

```{r,echo=FALSE}
ggplot(gplay, aes(x= Category, y= Reviews, fill = Type)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  ggtitle("Number Of App Reviews Based On Category and Type")
```

###Number of Apps Installs

```{r,echo=FALSE}
ggplot(gplay, aes(x= Category, y= Installs, fill = Type)) +
  geom_bar(position='dodge',stat='identity') +
  coord_flip() +
  ggtitle("Number Of App Installs Based On Category and Type")
```







